---
title: "The Causal Model and Notation"
engine: knitr
date-modified: last-modified
bibliography: references.bib
nocite: |
  @diaz2023nonparametric
---

{{< include macros.qmd >}}

Consider the set of random variables:

$$
\begin{align}
L &= f_L(U_L)\\
A &= f_A(L, U_A)\\
Y &= f_Y(A, L, U_Y)
\end{align}
$$

-   The functions $f$ are assumed fixed, but likely unknown.

-   $U = (U_L, U_A, U_Y)$ is a vector of exogenous, unmeasured random variables.

-   This system is referred to as a Non-parametric Structural Equation Model (NPSEM).

-   We will assume that this system of equations exists in nature and is responsible for generating our observed data.

::: {.callout-note appearance="simple"}
We will often refer to $L$ as confounders, $A$ as the treatment or exposure, and $Y$ as the outcome.
:::

------------------------------------------------------------------------

Take for example the following sample of data:

```{r echo=FALSE}
read.csv("data/williams_synthetic.csv") |>
  dplyr::select(sex, bmi, age, smoke, trt = A, event) |> 
  dplyr::sample_n(10) |> 
  DT::datatable() |> 
  DT::formatRound(columns = "age", digits = 0)
```

\
Let, $L$ = `sex`, `bmi`, `age`, and `smoke`; $A$ = `trt`; and $Y$ = `event`. When we say that we assume the previous NPSEM, we are positing that:

-   the variable `trt` was generated from an unknown function of the random variables `sex`, `bmi`, `age`, and `smoke`, as well as a set of unmeasured random variables not present in our data.

-   that the variable `event` was generated from an unknown function of the variables that generated `trt` in addition to `trt` itself.

Note that in some cases, we may know one or more of the functions $f$. For example, if our data came from a randomized clinical trial for $A$, then we are likely to know the true value of $f_A$.

------------------------------------------------------------------------

Central to how we will define causal effects is the concept of *counterfactual random variables.*

::: {.callout-important icon="false"}
## Counterfactual random variables

Hypothetical random variables that would have been observed, possibly contrary to fact, in an alternative world.
:::

For example, consider a scenario where we are interested in the value of $Y$ in a hypothetical situation where, instead of the variable $A$ being equal to its observed value, $A$ is set to some other value.

-   The value of $Y$ in this hypothetical situation is a counterfactual random variable.

-   let $\dd(A, L)$ be a function that returns a new value of $A$. **We will refer to the function** $\dd$ **as a *shift* function or a *general hypothetical intervention*.**

-   Denote the value of $Y$ in the hypothetical world where $A$ is set to the output of $\dd$ as $Y^{\dd}$.

Returning to the data example, imagine we are interested in the value of `event` if `trt` was replaced with the output of a function $\dd$ that always returns 1:

$$
\begin{align}
L &= f_L(U_L)\\
A &= \dd(A, L) = 1 \\
Y^{\dd} &= f_Y(1, L, U_Y)
\end{align}
$$

If we had the ability to collect data from this alternative NPSEM, the data may instead look like this:

```{r echo=FALSE}
read.csv("data/williams_synthetic.csv") |>
  dplyr::select(sex, bmi, age, smoke, trt = A, event) |> 
  dplyr::slice(1:10) |> 
  dplyr::mutate(trt = 1, event = 0) |> 
  DT::datatable() |> 
  DT::formatRound(columns = "age", digits = 0)
```

::: {.callout-note appearance="simple"}
Unfortunately, we are never able to collect data from this alternative world. This is called the *fundamental problem of causal inference.*
:::

------------------------------------------------------------------------

The previous NPSEM is the simplest causal model we will assume in this workshop. However, real data is often much more complex and may be characterized by:

-   time-varying variables

-   loss-to-follow-up

As such, we need to modify and introudce some additional notation:

| Symbol                                 | Definition                                                                                                                  |
|--------------------------|----------------------------------------------|
| $i$                                    | The index (i.e. a row in a dataset) of an observation from a data set with $n$ total units (i.e., the total number of rows) |
| $t$                                    | The index of time for a total number of time points $\tau$                                                                  |
| $L_t$                                  | Confounders at time $t$                                                                                                     |
| $A_t$                                  | A vector of intervention variables (i..e, treatment or exposure) at time $t$                                                |
| $Y$                                    | An outcome variable observed at time $\tau + 1$. Earlier measures of the outcome can be included in $L_t$.                  |
| $C_t$                                  | A indicator variable that a unit is *observed* at time $t+1$                                                                |
| $O_1, ..., O_n$                        | A sample of $n$ i.i.d observations with $O = (L_1, A_1, C_1, L_2, A_2, C_2, ..., L_\tau, A_\tau, C_\tau, Y)$                |
| $\bar{X}_t = (X_1, ..., X_t)$          | The history of a variable up until time $t$                                                                                 |
| $\underline{X}_t = (X_t, ..., X_\tau)$ | The future of a variable, including time $t$                                                                                |
| $H_t = (\bar{A}_{t-1}, \bar{L}_t)$     | The history of all variables up until just before $A_t$                                                                     |
| $\epsilon_t$                           | A randomizer                                                                                                                |
| $\dd_t(a_t, h_t, \epsilon_t)$          | A function that maps $A_t$, $H_t$, and $\epsilon_t$ to a new value of $A_t$                                                 |
