---
title: "Modified treatment policies"
engine: knitr
date-modified: last-modified
filters:
  - webr
webr:
  packages: ['lmtp', 'superlearner', 'rpart', 'gam', 'rlang']
  autoload-packages: true
  repos: 
    - https://nt-williams.r-universe.dev
bibliography: references.bib
nocite: |
  @diaz2023nonparametric, @haneuse2013estimation
---

{{< include macros.qmd >}}

```{webr-r}
#| context: setup
download.file("https://raw.githubusercontent.com/nt-williams/lmtp-workshop/main/data/crime_toy.csv", "crime.csv")
crime <- read.csv("crime.csv")
```

As we already discussed, a modified treatment policy is an intervention characterized by a dependence on the natural, or observed, value of treatment. Examples include:

-   Decreasing surgeries that last more than 4 hours by 1 hour (additive MTP)

-   Decreasing BMI by 2 units for individuals with BMI \> 30 (additive MTP)

-   Decreasing AQI by 20% (multiplicative MTP).

## Synthetic Neighborhood Community Violence Data

In some current work, we are interested in estimating the effect of decreasing the level of community violence (violent crime rate) on risk of preterm birth. Research suggests community violence may impact infant health through physiological responses to stress, worsening maternal mental health, and unhealthy coping behaviors. Community violence is a highly endogenous exposure, which, motivates the use of such a shift intervention.[^1]

[^1]: Thank you to Chris Dharma, who simulated this data!

```{r echo=FALSE}
crime <- read.csv("data/crime_toy.csv")
dplyr::select(crime, -community_code) |> 
  gtsummary::tbl_summary()
```

## Problem 1

Write a shift function (named `d_mtp`) for a modified treatment policy where the natural value of neighborhood violence is decreased by 50%. Using the `lmtp` package, estimate the population mean effect of this intervention on preterm birth. Make sure to use the `id` argument to account for community level clustering in the estimation of standard errors. To save computation time, don't use cross-fitting.

::: {.callout-warning icon="false"}
## Practice

```{webr-r}
A <- "neighborhood_violence"
W <- names(crime)[1:12]
Y <- "preterm_birth"
id <- "community_code"

learners <- c("SL.mean", 
              "SL.glm", 
              "SL.rpart", 
              "SL.gam")

d <- # Write your shift function here!

fit_d <- # Write your estimation code here!
```

<details>

<summary style="color: black; font-weight: bold;">

Solution

</summary>

```{r eval=FALSE}
d_mtp <- function(data, a) {
  data[[a]]*0.5
}

fit_mtp <- lmtp_tmle(
  data = crime, 
  trt = A, 
  outcome = Y, 
  baseline = W,
  id = id,
  shift = d_mtp, 
  mtp = TRUE,
  folds = 1, 
  learners_trt = learners, 
  learners_outcome = learners
)
```

</details>
:::

Did you get this error?

```{r eval=FALSE}
Warning: Detected decimalish `trt` values and `mtp = FALSE`. Consider setting `mtp =
TRUE` if getting errors.
Loading required package: nnls
Error: object 'fit' not found
```

::: {.callout-tip appearance="simple"}
Make sure to set `mtp = TRUE` if you're intervention is a modified treatment policy!
:::

## Problem 2

Let's compare the effect of the intervention on reducing neighborhood violence on preterm birth to the observed preterm birth rate. Use the `lmtp_contrast()` function.

::: {.callout-warning icon="false"}
## Practice

```{webr-r}
# Write your code here!
```

<details>

<summary style="color: black; font-weight: bold;">

Solution

</summary>

```{r eval=FALSE}
lmtp_contrast(fit_mtp, ref = mean(crime$preterm_birth))
```

</details>
:::

With `lmtp` we found that under a hypothetical intervention that reduces neighborhood violence by 50%, preterm birth would decrease by approximately 4.4 percentange points.
