---
title: "Static Interventions"
engine: knitr
date-modified: last-modified
filters:
  - webr
webr:
  packages: ['lmtp', 'superlearner']
  autoload-packages: true
  repos: 
    - https://nt-williams.r-universe.dev
    - https://imbs-hl.r-universe.dev
bibliography: references.bib
nocite: |
  @goyal2020clinical, @williams2022optimising
tbl-cap-location: top
---

{{< include macros.qmd >}}

```{webr-r}
#| context: setup
download.file("https://raw.githubusercontent.com/nt-williams/lmtp-workshop/main/data/williams_synthetic.csv", "williams_synthetic.csv")
covid <- read.csv("williams_synthetic.csv")
```

### Synthetic COVID RCT Data

For our first example, we'll use a synthetic dataset of $n = 1000$ patients from a hypothetical clinical trial for a treatment to decrease intubation and death among patients hospitalized with COVID-19.

```{r echo=FALSE}
read.csv("data/williams_synthetic.csv") |> 
  DT::datatable(class = "compact", 
                options = list(
                  searching = FALSE, 
                  lengthChange = FALSE, 
                  ordering = FALSE, 
                  info = FALSE, 
                  columnDefs = list(
                    list(targets = 0, visible = FALSE),
                    list(className = 'dt-left', targets = "_all")  # apply to all columns
                  )
                )) |> 
  DT::formatRound(columns = "age", digits = 0)
```

\

-   The data is based on a database of over 1,500 patients hospitalized at Weill Cornell Medicine New York Presbyterian Hospital prior to 15 May 2020 with COVID-19 confirmed through PCR.

-   To replicate a two-arm randomized clinical trial (RCT), we've simulated a hypothetical treatment variable (`A`) that is randomly assigned for each observation with probability 0.5.

-   The outcome of interest (`event`) is intubation or death by 15-days post-hospitalization.

-   Treatment is associated with increased survival.

-   Baseline covariates include: age, sex, BMI, smoking status, whether the patient required supplemental oxygen within 3 hours of presenting to the emergency department, number of comorbidities, number of relevant symptoms, presence of bilateral infiltrates on chest X-ray, dyspnea, and hypertension.

```{r echo=FALSE}
covid <- read.csv("data/williams_synthetic.csv")
dplyr::select(covid, -days, -id) |> 
  gtsummary::tbl_summary(by = A)
```

### Population mean outcomes

Our goal is to estimate the average treatment effect (ATE) of the hypothetical treatment on intubation or death:

$$
\begin{align}
\theta &= \E(Y^{A=1} - Y^{A=0}) \\
&= \textcolor{blue}{\E(Y^{A=1})} - \textcolor{red}{\E(Y^{A=0})}
\end{align}
$$

Notice that the ATE is composed of two parameters:

1.  the proportion of patients who receive intubation or die in a hypothetical world where all patients receive treatment ($\textcolor{blue}{\E(Y^{A=1})}$), and

2.  the proportion of patients who receive intubation or die in a hypothetical world where no patients receive treatment ($\textcolor{red}{\E(Y^{A=0})}$).

We refer to these parameters as *population mean outcomes.* Let's rewrite the target parameter as a function of interventions using the framework we previously discussed. Let $\dd_1(a, h) = 1$ and $\dd_0(a, h) = 0$. Then,

$$
\theta = \textcolor{blue}{\E(Y^{\dd_1})} - \textcolor{red}{\E(Y^{\dd_0})}.
$$

Remember, we refer to these interventions as static because they return the same value regardless of their input.

#### Writing shift functions

So, how do we translate the functions $\dd_1$ and $\dd_0$ into R code to be used with `lmtp`?

::: {.callout-note appearance="simple"}
Policies can be specified using one of two arguments in `lmtp` estimators: `shift` or `shifted`.
:::

If using `shift`, we supply a two-argument function of the form

```{r eval=FALSE}
d <- function(data, trt) {
  # Insert code here
}
```

-   The first argument should correspond to a dataset containing all the variables in $O$.

-   The second argument should expect a string corresponding to the variable(s) $A_t$ in $O$.

-   This function should return a size $n$ vector with the same class as $A_t$ modified according to $d_t$.

For $\dd_1$ this function would be:\

```{webr-r}
#| context: interactive
#| autorun: true
d1 <- function(data, trt) {
  rep(1, nrow(data))
}
```

If we apply this function to our data we are returned a vector of 1s with the same length as the number of rows in the observed data.

```{webr-r}
d1(covid, "A") |> head()
d1(covid, "A") |> length()
```

### Problem 1

::: {.callout-tip  appearance="minimal"}
**Question**: Which estimator should we use to estimate these parameters?

```{=html}
<details>
<summary style="color: black; font-weight: bold;">Answer</summary>
<p>Because this is a non-time-varying study, we should use TMLE.</p>
</details>
```
:::

------------------------------------------------------------------------

Let's now estimate $\E(Y^{\dd_1})$ using `lmtp`.

```{webr-r}
A <- "A"
Y <- "event"
W <- c("age", "sex", "bmi", "smoke", "o2", 
       "num_comorbid", "num_symptoms", "bilat", "dyspnea", "hyper")

set.seed(34465)

treat <- lmtp_tmle(
  data = covid, 
  trt = "A", 
  outcome = "event", 
  baseline = W, 
  outcome_type = "binomial", 
  shift = d1, 
  folds = 1, 
  learners_trt = "SL.glm", 
  learners_outcome = "SL.glm"
)

print(treat)
```

Instead of specifying an intervention using `shift`, we could instead supply a modified version of `data` to the `shifted` argument where the variables $A_t$ are replaced with $A^d_t$.\

```{webr-r}
tmp <- covid
tmp$A <- d1(covid, "A")

head(tmp)

set.seed(34465)

treat <- lmtp_tmle(
  data = covid, 
  trt = "A", 
  outcome = "event", 
  baseline = W, 
  outcome_type = "binomial", 
  shifted = tmp, 
  folds = 1, 
  learners_trt = "SL.glm", 
  learners_outcome = "SL.glm"
)

print(treat)
```

### Problem 2

::: {.callout-tip  appearance="minimal"}
**Question**: How can we interpret this result?

```{=html}
<details>
<summary style="color: black; font-weight: bold;">Answer</summary>
<p>In a hypothetical world where all patients received treatment, the expected proportion of patients who receive invasive mechanical ventilation by day 15 of hospitalization is 0.2722 (95% CI: 0.24 to 0.30).</p>
</details>
```
:::

### `lmtp` objects

A call to an `lmtp` estimator returns a list of class `lmtp`.

| Value          | Description                                                                                                                  |
|----------------|------------------------------------------------------------------------------------------------------------------------------|
| estimator      | The estimator used.                                                                                                          |
| theta          | The estimated population LMTP estimand.                                                                                      |
| standard_error | The estimated standard error of the LMTP estimand.                                                                           |
| low            | Lower bound of the 95% confidence interval of the LMTP estimand.                                                             |
| high           | Upper bound of the 95% confidence interval of the LMTP estimand.                                                             |
| eif            | The estimated, un-centered, influence function of the estimate.                                                              |
| shift          | The shift function specifying the treatment policy of interest.                                                              |
| outcome_reg    | An $n \times \tau + 1$ matrix of outcome regression predictions. The mean of the first column is used for calculating theta. |
| density_ratios | An $n \times \tau$ matrix of the estimated, non-cumulative, density ratios.                                                  |
| fits_m         | A list the same length as `folds`, containing the fits at each time-point for each fold for the outcome regression.          |
| fits_r         | A list the same length as `folds`, containing the fits at each time-point for each fold of density ratio estimation.         |
| outcome_type   | The outcome variable type.                                                                                                   |

: Values returned in an `lmtp` object.

Let's inspect some of these values.

```{webr-r}
hist(treat$density_ratios[, 1], 
     main = "Histogram of density ratios", 
     xlab = "Density ratio")
```

### Problem 3

Write a function to estimate the population mean outcome if no patients received treatment. Assign the result to an object named `dont_treat`.

::: {.callout-tip  appearance="minimal"}

```{webr-r}
d0 <- # Write your shift function here!
dont_treat <- # Type your code here!
```

<details>

<summary style="color: black; font-weight: bold;">

Solution

</summary>

```{r eval=FALSE}
d0 <- function(data, trt) {
  rep(0, nrow(data))
}

dont_treat <- lmtp_tmle(
  data = covid, 
  trt = "A", 
  outcome = "event", 
  baseline = W, 
  outcome_type = "binomial", 
  shift = d0, 
  folds = 1, 
  learners_trt = "SL.glm", 
  learners_outcome = "SL.glm"
)

print(dont_treat)
```

</details>
:::

::: {.callout-note appearance="simple"}
`lmtp` already implements `d1` and `d0` as `static_binary_on()` and `static_binary_off()`!
:::

### Average treatment effect

With estimates of $\E(Y^{\dd_1})$ and $\E(Y^{\dd_0})$ we can now calculate the ATE. To do so, we'll use the function `lmtp_contrast()`.

```{webr-r}
lmtp_contrast(treat, ref = dont_treat)
```

In addition to the ATE, we can also calculate the causal risk ratio and the causal odds ratio.

```{webr-r}
lmtp_contrast(treat, ref = dont_treat, type = "rr")
lmtp_contrast(treat, ref = dont_treat, type = "or")
```

Confirming what we already know, we've estimated the treatment as being associated with decreased intubation or death.
